Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2018-09-28 Response_to_Reviewers.tex

    \item More generally concerning the choice of these two models, we aim at helping the readers to distance themselves from the case studies and focus on the genericity of the approach. This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs. This is again an argument in favor of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.

	% 3. I am not sure why the Sugarscape model has been chosen given that the paper seems to be focused on cities.
	\item Furthermore concerning the choice of the Sugarscape model, we think it is a good example of a stylized model of a spatialized socio-economic system, and according to (Batty, 2005) a good metaphor of an urban system. Our choice was accordingly clarified in the paper.

	% 4. Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”
	\item We agree that the term of meta-parameters may have been confusing, and following the suggestion we replaced it with ``generator parameters''.

	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
    \item There was a mistake in the writing, as we never use the median but always the mean, which as you recall is consistent with the choice of distance we made. This was corrected in the paper.

	% 6. Line 232: comparing phase diagrams is related to the topic of comparing patterns. It would make sense to mention the methods that are usually used to compare patterns
	\item Some general references on comparing patterns were added.

	% 7. Line 238: Please explain why this particular index was chosen and discuss its properties. Connect the discussion to other indices like L1, and Linf norms, etc. Would we get qualitatively different results from different indices? 
	\item The approach proposed in the paper is indeed rather exploratory. This explains why we have chosen to focus on a single index (L2) as this index is relatively well-known in the community. Previous exploratory analyses have shown that the results were rather stable when using the Monge method. Adapting the mathematical aggregator to the different indices would constitute an interesting perspective, but is out of the scope of the paper which already focuses on two models and proposes a new framework for analysing the impact of initial pattern parameters. We added a supplementary material showing the distances obtained with other distances, which are similar to the results in main text.

@@

	% The most interesting part of the research to me, is not the way the density layers are generated but the use of the phase diagrams, defined as a vector of final aggregated model outputs. As these are vectors, they are not so different from time series and many techniques exist to compare vectors in this domain. I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.
	\item The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distances do not change much qualitative results (analysis added in supplementary material) and with more literature for possible extensions.

	% Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
	\item The functionalities of OpenMOLE we used were detailed in the article; we did not use genetic algorithms but sampling methods and high performance computing environments.

	% In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions. Actually this came as a surprise to me. I had the impression that all phase diagrams would be evaluated against all other diagrams. Perhaps this should be explained earlier in the paper.
	\item Concerning the measure of the distance of all phase diagram to one reference phase diagram, this choice was made as we preferred having a local sensitivity index for which the distribution can be studied, rather than a global completely integrated sensitivity index (similar to Saltelli or Morris indexes for example). This was explained better in the paper.

@@

	% Geosimulation is introduced but not defined or referenced.  This should be clarified as the term has several meanings etc.
	\item For the definition of geosimulation, we meant the one coined by (Benenson and Torrens, 2004), this was made explicit and the reference was added.

	% In places the English along with the overall argument could be better articulated. For example, on page 1 you write: “We think simulation can become a very good tool to achieve this,”  Could this be rewritten or better argued. You are not the only one to believe this but you don’t show this in your paper.
	\item We have worked towards improving the English along with the overall argument.

	% Page 4: you write “Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time” This makes little sense. What it should be is “Generally, sensitivity analyses have focused either on the spatial context or the heterogeneity of space” or something like that. Basically you need the “or”.
	\item Thank you for the suggestion on page 4. We have modified the paragraph accordingly.

	% You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?  
	\item We do not have knowledge of examples in which the comparison of patterns or fields is called ``phase diagram comparison'', and we neither have knowledge of such work in physics for example, from which the term originated. There is naturally all the literature we cited on similar issues in other domains, and the extension of the discussion gives elements of answer to this point.

### 2018-09-28 paper.tex
Update on Overleaf.

\documentclass[preprint,5p,times,twocolumn,authoryear]{elsarticle}

@@

In order to test the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}. Following \citet{Gauvinetal2009}, we define a phase diagram as the vector of final aggregated model outputs considered as a function of model parameters. We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment). A solution is to use systematic quantitative procedures to compare them \added{to a reference case}. Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median \added{?} of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\

@@

where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given \replaced{generator}{meta-} parameters $\vec{\alpha}_i$. $d$ is a functional distance that we take simply as Euclidean distance. The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we \replaced{average}{sum} these relative distances over the components.

@@

@@

We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default initial spatial condition setup (a bi-centric symmetrical non toroidal configuration) using the measure defined in equation~\ref{eq:phase-distance}. For each density grid, we obtain the average squared distance between corresponding points of the phase diagrams, i.e. the median\added{?} value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated). This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry). Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.

@@

@@

\added{For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\mathbb{E}{X}}/\hat{\sigma}(X)$ with standard estimators for average and standard deviation. The summary statistics computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.}

%%%%%%%%%%%%%%
\begin{table*}
\caption{\added{\textbf{Summary statistics of Sharpe ratio estimated on repetitions for each parameter point.}}\label{tab:summarysharpes}}
\centering
	\begin{tabular}{|ccccccc|}
	\hline
Model/Indicator & Min. & 1st Qu. &  Median &  Mean & 3rd Qu. & Max. \\\hline
Sugarscape/Gini & 2.343 & 14.413 & 20.993 & 20.098 & 26.208 & 49.808\\\hline 
Schelling/Dissimilarity & 1.053 & 12.368 & 19.434 & 25.268 & 24.328 & 468.563\\
Schelling/Entropy & 0.5458 & 8.2591 & 14.5112 & 17.0100 & 17.9372 & 244.3230\\
Schelling/Moran & 0.0001 & 0.5554 & 0.7782 & 3.2724 & 3.8979 & 121.6829\\\hline
    \end{tabular}
\end{table*}
%%%%%%%%%%%%%%

@@

\added{We describe here the tests done with other distances to compare phase diagrams. We tested normalized Minkovski distances, defined by $d(x,y) = \left(\frac{1}{N}\cdot \sum_i \left|x_i - y_i\right|^{q}\right)^{\frac{1}{q}}$, for varying values of $q$ from $q = 1$ (Manhattan distance) to $q = 10$, including $q = 2$ (Euclidian distance) which is used in main text.}

@@

\begin{table*}
	\caption{\added{\textbf{Summary statistics of different distances for schelling.}}}
	\centering
	\begin{tabular}{|ccccccc|}
	\hline
    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
        1 & 0.00000 & 0.02838 & 0.06263 & 0.10608 & 0.16818 & 0.37018 \\
        2 & 0.0000 & 0.1520 & 0.2517 & 0.3107 & 0.4359 & 0.8155 \\
        3 & 0.0000 & 0.3709 & 0.5133 & 0.5860 & 0.7435 & 1.2930 \\
        10 & 0.000 & 2.083 & 2.431 & 2.380 & 2.713 & 3.664 \\\hline
    \end{tabular}
\end{table*}

\begin{table*}
	\caption{\added{\textbf{Summary statistics of different distances for sugarscape.}}}
	\centering
	\begin{tabular}{|ccccccc|}
	\hline
    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
        1 & 0.07184 & 0.13996 & 0.39816 & 0.42452 & 0.64751 & 1.13169\\
        2 & 0.08909 & 0.19794 & 1.52272 & 1.29594 & 2.16371 & 2.98273\\
        3 & 0.1025 & 0.2288 & 3.1495 & 2.5074 & 4.2718 & 5.5378\\
        10 & 0.1540 & 0.4466 & 9.8757 & 7.6246 & 13.0550 & 16.8001\\\hline
    \end{tabular}
\end{table*}

@@

\added{We give below the pseudo-code for the implementations we used of both sugarscape and schelling models. We recall that source code is openly available at} \url{https://github.com/AnonymousAuthor2/SpaceMatters}.

>>>>

\documentclass[preprint,5p,times,twocolumn,authoryear,draft]{elsarticle}

@@

In order to test the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}. Following \citet{Gauvinetal2009}, we define a phase diagram as the vector of final aggregated model outputs considered as a function of model parameters. We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment). A solution is to use systematic quantitative procedures to compare them \added{to a reference case}. Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the \replaced{mean}{median} of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\

@@

where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given \replaced{generator}{meta-} parameters $\vec{\alpha}_i$. $d$ is a functional distance that we take simply as Euclidean distance. The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$. For a multi-dimensional phase diagram, we \replaced{average}{sum} these relative distances over the components. \added{Given a set of phase diagram to be compared, we will study the distribution of this distance to an arbitrary phase diagram for all diagrams, rather than an aggregated measure which would be similar to global sensitivity methods \citep{saltelli2008global}.}

@@

\added{As detailed in supplementary material~\ref{app:convergence}, more repetitions are needed for Schelling indicators than for Sugarscape, in order to obtain a similar relative confidence in the estimation of averages. We ran for this reason a different number of replications for each model.}

@@

We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default initial spatial condition setup (a bi-centric symmetrical non toroidal configuration) using the measure defined in equation~\ref{eq:phase-distance}. For each density grid, we obtain the average squared distance between corresponding points of the phase diagrams, i.e. the \replaced{mean}{median} value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated). This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry). Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability. \added{We tested the sensitivity to the type of distance, using a Minkovski distance with a varying exponent. The results are presented in supplementary material~\ref{app:distances}, and show a similar sensitivity to spatial initial conditions.}

@@

\label{app:convergence}

@@

\added{For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\mathbb{E}{X}}/\hat{\sigma}(X)$ with standard estimators for average and standard deviation. The summary statistics of this ratio computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}. Under the assumption of a normal distribution, the width of the confidence interval at level $c$ is given by $\left|\mu_+ - \mu_-\right| = 4\cdot\sigma \sigma z_{c} \sqrt{n}$ where $\sigma$ is the standard deviation, $z_{c}$ is the quantile at which $c$ is attained by the cumulative distribution, which is around $1.96$ for a 95\% confidence interval. This means that to obtain a confidence interval of width $\kappa \cdot \sigma$, one needs a number $n \simeq (8 / \kappa )^2$ of repetitions. This gives 64 repetitions for $\kappa = 1$. As the Sharpes ratios are in general smaller for Schelling indicators than for Sugarscape, we take $n = 50$ for Sugarscape and $n = 100$ for Schelling to have s similar confidence in estimations.}

%%%%%%%%%%%%%%
\begin{table*}
\caption{\added{\textbf{Summary statistics of Sharpe ratio estimated on repetitions for each parameter point.}}\label{tab:summarysharpes}}
\centering
	\begin{tabular}{|c|cccccc|}
	\hline
Model/Indicator & Min. & 1st Qu. &  Median &  Mean & 3rd Qu. & Max. \\\hline
Sugarscape/Gini & 2.343 & 14.413 & 20.993 & 20.098 & 26.208 & 49.808\\\hline 
Schelling/Dissimilarity & 1.053 & 12.368 & 19.434 & 25.268 & 24.328 & 468.563\\
Schelling/Entropy & 0.5458 & 8.2591 & 14.5112 & 17.0100 & 17.9372 & 244.3230\\
Schelling/Moran & 0.0001 & 0.5554 & 0.7782 & 3.2724 & 3.8979 & 121.6829\\\hline
    \end{tabular}
\end{table*}
%%%%%%%%%%%%%%

@@

\label{app:distances}

\added{We describe here the tests done with other distances to compare phase diagrams. We tested normalized Minkovski distances, defined by $d(x,y) = \left(\frac{1}{N}\cdot \sum_i \left|x_i - y_i\right|^{q}\right)^{\frac{1}{q}}$, for varying values of $q$ from $q = 1$ (Manhattan distance) to $q = 10$, including $q = 2$ (Euclidian distance) which is used in main text. The Table~\ref{tab:schellingdistances} gives the summary statistics of each distance computed on all initial configurations for the Schelling model. The Table~\ref{tab:sugarscapedistances} gives the same statistics for the sugarscape model. We naturally obtain smaller difference with the Manhattan distance but which remain significant (averages of 10\% for the Schelling model and 40 \% for Sugarscape), and variabilities with higher values of the Minkovski exponent are much higher. This results confirm the high variability observed in main text with the Euclidian distance.}

@@

\begin{table*}
	\caption{\added{\textbf{Summary statistics of different distances for schelling.}}}\label{tab:schellingdistances}
	\centering
	\begin{tabular}{|ccccccc|}
	\hline
    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
        1 & 0.00000 & 0.02838 & 0.06263 & 0.10608 & 0.16818 & 0.37018 \\
        2 & 0.0000 & 0.1520 & 0.2517 & 0.3107 & 0.4359 & 0.8155 \\
        3 & 0.0000 & 0.3709 & 0.5133 & 0.5860 & 0.7435 & 1.2930 \\
        10 & 0.000 & 2.083 & 2.431 & 2.380 & 2.713 & 3.664 \\\hline
    \end{tabular}
\end{table*}

\begin{table*}
	\caption{\added{\textbf{Summary statistics of different distances for sugarscape.}}}\label{tab:sugarscapedistances}
	\centering
	\begin{tabular}{|ccccccc|}
	\hline
    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
        1 & 0.07184 & 0.13996 & 0.39816 & 0.42452 & 0.64751 & 1.13169\\
        2 & 0.08909 & 0.19794 & 1.52272 & 1.29594 & 2.16371 & 2.98273\\
        3 & 0.1025 & 0.2288 & 3.1495 & 2.5074 & 4.2718 & 5.5378\\
        10 & 0.1540 & 0.4466 & 9.8757 & 7.6246 & 13.0550 & 16.8001\\\hline
    \end{tabular}
\end{table*}

@@

\added{We give below the pseudo-code for the implementations we used of both sugarscape and schelling models. We recall that source code is openly available at} \url{https://github.com/AnonymousAuthor2/SpaceMatters}. \added{The pseudo-code is in the style of NetLogo code, which is already easily readable.}
